How to build an homogeneous language for heterogeneous platforms

### 2013-05-30 papers/gpce2013/gpce2013.tex
abstract

\begin{abstract}
Writing code partitioned between heterogeneous client-side and server-side makes the engineering process of Web
applications tedious. We use a deep embedded DSL to define a common language abstracting over platform differences
for the developers point of view, but generating efficient code leveraging the target platforms characteristics and
giving them control on the amount of code produced on client-side.
\end{abstract}

@@

The more interactive the application is, the more logic needs to be duplicated between the server-side
and the client-side (explain why?), and the higher is the amount of client-side code. Developers use libraries and
frameworks to get higher-level abstractions on client-side, making their code easier to reason about and to
maintain. Unfortunately, JavaScript -- which is currently the most supported the action language on Web clients
-- has no static linking system allowing to pick just the functions actually used by a program, forcing clients to
download the whole library even though only a few functions are actually used.

@@

\begin{itemize}
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
 \item Static linking.
\end{itemize}

@@

Problem: interoperability.

@@

The application was initially written using jQuery. We rewrote it using several technologies for the client-side
part: plain JavaScript (without third-party library), js-scala, GWT and SharpKit.

@@

Our goal is to evalute the level of abstraction provided by each solution and their performances. We took the
number of lines of code as a measure of the level of abstraction and we wrote micro-benchmarks to evaluate the
runtime performances of a solution. We also measured the size of the data sent to the client-side. We also measured
the ability to share code between client and server sides.

>>>>

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between client and server code. Our DSL gives a
performance / expressiveness ratio about two times higher than other approaches.
\end{abstract}

@@

The more interactive the application is, the more logic needs to be duplicated between the server-side
and the client-side (explain why?), and the higher is the amount of client-side code. Developers use libraries and
frameworks to get high-level abstractions on client-side, making their code easier to reason about and to
maintain. Unfortunately, JavaScript -- which is currently the most supported the action language on Web clients
-- has no static linking system allowing to pick just the functions actually used by a program, forcing clients to
download the whole library even though only a few functions are actually used.

@@

\begin{itemize}
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
 \item Static linking?
\end{itemize}

@@

Another approach consists in defining several small languages, each one being specific to a given set of concerns,
and then to combine all the source artifacts written with these language into one executable program. This approach
makes it difficult to make the separate languages inter-operate with each other.

@@

The application was initially written using jQuery. We rewrote it using several technologies for the client-side
part: plain JavaScript (without third-party library), js-scala, GWT and HaXe.

\paragraph*{Vanilla JavaScript} Low-level code.

\paragraph*{jQuery} High-level code.

\paragraph*{js-scala} High-level code. HTML fragment definition reused between server and client sides.

\paragraph*{HaXe} Low-level code.

\paragraph*{GWT} High-level code?

@@

Our goal is to evalute the level of abstraction provided by each solution and their performances. We took the
number of lines of code as a measure of the level of abstraction. We also measured the size of the data sent to the
client-side. We also measured the ability to share code between client and server sides.

Rather than doing a micro-benchmark focusing on just one abstraction, we performed a global benchmark that is more
likely to reflect the real performances of the application: we simulated user actions on a Web page (2000 clicks on
buttons, triggering a dynamic update of the page and involving the use of the Option monad, the selectors API and
the HTML fragment definition API) and measured the time it took complete them. We only measured the execution time of
client-side code execution. The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google
Chrome v27 Web browser.

\begin{figure}
\label{benchmark}
\centering
\includegraphics[width=8cm]{chooze.png}
\caption{Benchmarks on a real application}
\end{figure}
